Type theories as quotient inductive-recursive types

### 2025-01-31 tex/main.tex
draft intro

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen and T.-C.~Tsai}
\Copyright{Liang-Ting Chen and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Fredrik Nordvall Forsberg, Hsiang-Shang Ko, and Meven Lennon-Bertrand.}%optional

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}

@@

\section{Introduction} \label{sec:intro}
\LT{%
Points to elaborate:
\begin{enumerate}
  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).

>>>>

\title{Type theories as quotient inductive-recursive types}
\author{Liang-Ting Chen\footnote{Corresponding author; authors are listed in alphabetical order.}}{Institute of Information Science, Academia Sinica, Taiwan \and \url{http://l-tchen.github.io}}{ltchen@iis.sinica.edu.tw}{https://orcid.org/0000-0002-3250-1331}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 114-2222-E-001-001-MY3.}
\author{Fredrik Nordvall Forsberg}{Department of Computer and Information Sciences, University of Strathclyde, UK \and \url{https://fredriknf.com}}{fredrik.nordvall-forsberg@strath.ac.uk}{https://orcid.org/0000-0001-6157-9288}{Supported by the Engineering and Physical Sciences Research Council [EP/Y000455/1].}
\author{Tzu-Chun Tsai}{Institute of Information Science, Academia Sinica, Taiwan}{gene0905@icloud.com}{}{Supported by the National Science and Technology Council of Taiwan under grant NSTC 112-2221-E-001-003-MY3.}
\authorrunning{L.-T.~Chen, F.~Nordvall Forsberg, and T.-C.~Tsai}
\Copyright{Liang-Ting Chen, Fredrik Nordvall Forsberg, and Tzu-Chun Tsai}
\ccsdesc[500]{Theory of computation~Type theory}
\keywords{inductive-inductive types, quotient inductive types, inductive-recursive types, substitution calculus, category with families}
\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

\supplement{The formal development is hosted at the GitHub repository: \url{https://github.com/genetsai95/DTT-QIIRT}.}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional
\acknowledgements{We'd like to thank Hsiang-Shang Ko and Meven Lennon-Bertrand.}%optional

@@

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{YYY}
\EventNoEds{1}
\EventLongTitle{10th International Conference on Formal Structures for Computation and Deduction (FSCD 2025)}
\EventShortTitle{FSCD 2025}
\EventAcronym{FSCD}
\EventYear{2025}
\EventDate{July 14--20, 2025}
\EventLocation{Birmingham, United Kingdom}
\EventLogo{}
\SeriesVolume{XX}
\ArticleNo{XX}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage{manfnt}
\newcommand{\danger}{\marginpar[\hfill\dbend]{\dbend\hfill}}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},#1]{#2}}
\newcommand{\Gene}[2][]{\todo[inline,author={Gene},caption={},#1]{#2}}
\newcommand{\Fredrik}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

@@

\section{Introduction} \label{sec:intro}

Metaprogramming and mechanised metatheory in a type-theoretic proof assistant relies on having an internal representation of type theory within type theory itself.
%
In the spirit of type theory, we would like to represent well scoped and well typed terms only --- this also has technical advantages, such as theorem statements not having to deal with impossible cases, or separately having to prove wellformedness lemmas, etc.
%
Doing this effectively seems to require several sophisticated features of the host type theory, most notably advanced inductive definitions.
%
For example, inductive families~\cite{Dybjer1994} can be used to represent the simply typed lambda calculus~\cite{Altenkirch1999}, while some kind of simultaneous definitions such as inductive-recursive~\cite{Dybjer2000} or inductive-inductive~\cite{NordvallForsbergPhD} definitions seem to be required for dependent type theory~\cite{Danielsson2006,Chapman2009}.
%
This is not just a matter of the pure strength of the host theory, but also expressivity and convenience: while it is known how to reduce inductive-inductive definitions to inductive families in principle~\cite{Kaposi2000}, the resulting development would be considerable harder to work with, and most likely would have exhausted the resources of both the proof assistant and its user.

Altenkirch and Kaposi~\cite{Altenkirch2016a} showed that by employing \emph{quotient inductive-inductive types}~\cite{NordvallForsberg2018}, the internal representation of well typed terms could be significantly simplified, since equations such as $\beta$- and $\eta$-equality can be represented by the identity type thanks to quotient constructors.
%
In this paper, we show that the development can be further streamlined by combining quotient inductive-inductive types and inductive-recursive types.
%
The basic idea, following a suggestion by Kaposi~\cite{Kaposi2023}, is to replace the inductively defined action of a substitution on a type with a recursive definition, which computes.
%
As a result, more equations will hold strictly, and we can (at least partially) avoid the so-called ``transport hell'' that arises when constructions need to be transported across equality constructors.
%
This is a general phenomenon when working with quotient inductive-inductive types: to make use of equalities in later constructors and when using the eliminator for the type, the equalities need to be transported across.
%
Any method that reduces the friction this introduces seem worth exploring.

%However, we found that using quotient inductive-inductive-recursive types alone were not sufficient for a smooth development.
%%
%%Not only do we want the action of substitutions on types to compute definitionally, but we also want as many equations to hold strictly as possible:
%%
%%on the one hand, we want substitutions to distribute over type formers, but on the other hand we want substitutions to satisfy composition and unit laws of the category of context and substitutions up to definitional equality.
%%
%Instead, we also needed to make use of functions defined by overlapping patterns.
%%
%In a proof assistant such as Agda, this can be emulated by Agda's experimental support for rewrite rules, but with an additional proof burden on the user to show (informally) that their use is justified meta-theoretically.
%%
%To get some help with this task by Agda's confluence checker, it is often necessary to needlessly complicate a definition to convince the checker that the proposed rewrite rules are indeed confluent.

Our development also makes use of another feature not usually found in standard presentations of type theory: recursive pattern matching definitions with overlapping patterns.
%
This seems to be needed in order to again avoid transport hell in the form of equations that do not reduce due to the standard translation of pattern matching to case trees.
%
We emphasise that, following a fine tradition of previous attempts to represent type theory in type theory~\cite{Danielsson2006,Chapman2009,Altenkirch2016a}, we are making use of type-theoretic features that are so far not well understood or foundationally well motivated. Nevertheless, Agda can be convinced to accept our development, using some experimental flags. Indeed, we also see this paper as motivation to study quotient inductive-inductive-recursive types and overlapping patterns further.

\Fredrik{I'm not sure if I have included all points here. I tried to stay more positive re overlapping patterns.}
\LT{%
Points to elaborate:
\begin{enumerate}
  \item QIITs are littered with transports, causing too much pains while formalising a type theory.
  \item Rewrite rules are popular to make structural rules definitional but requires efforts to justify meta-theoretic properties in general.
  \item Further, the confluence check adopted by \Agda is based on complete development and requires one-step parallel reductions to have a confluent term, making a definition more complicated than necessary.
  \item Moreover, the use of rewrite rules is typically \emph{not} presented and discussed in literature. We are ignoring the elephant in the room.
  \item Instead, definitions by overlapping patterns (in theory) requires strong normalisation and local confluence, making definitions simpler to design.
  \item The rules for type substitutions are \emph{structural} shared with other type theories based on parallel substitution (in particular cwf).
